---
title: 'Practical: exploring RNA-Seq counts'
author: "Hugo Varet, Julie Aubert and Jacques van Helden"
date: '`r Sys.Date()`'
output:
  slidy_presentation:
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
---

```{r include=FALSE}
options(width=300)
knitr::opts_chunk$set(fig.align="center", cache=FALSE)
```

# Introduction

## Context

 - Study of 48 WT yeast samples vs 48 Snf2 (KO) samples: Gierliński et al. *Statistical models for RNA-seq data derived from a two-condition 48-replicate experiment*, Bioinformatics, 2015.

 - RNA-Seq reads have been cleaned, mapped and counted to generated a count data matrix containing 7126.

## Loading a data table

**R** enables to download data directly from the Web. Load counts per gene for each sample of biological condition (WT and Snf2).

```{r}
## Load the files content in an R data.frame
counts <- read.table("http://jvanheld.github.io/stats_avec_RStudio_EBA/practicals/yeast_2x48_replicates/data/counts.txt", sep="\t", header=TRUE, row.names=1)

expDesign <- read.table("http://jvanheld.github.io/stats_avec_RStudio_EBA/practicals/yeast_2x48_replicates/data/expDesign.txt", sep="\t", header=TRUE)

## look at the beginning of the counts and design table:
print(counts[1:6,1:5])
print(expDesign[1:5,])

## dimension
dim(counts)
dim(expDesign)
```

## Checking the content of the count tables

```{r eval=FALSE}
View(counts)
View(expDesign)

print(expDesign$strain)
expDesign$strain <- relevel(expDesign$strain, "WT")
print(expDesign$strain)
```

## Basic description of the data: number of reads per sample

```{r}
barplot(colSums(counts)/1000000, main="Total number of reads per sample (million)")
```

## Basic description of the data: percentage of null counts per sample

```{r}
prop.null <- apply(counts, 2, function(x) 100*mean(x==0))
print(prop.null)
barplot(prop.null, main="Percentage of null counts per sample")
```

# Differential analysis using DESeq2

## DESeq2 main object: dds

```{r message=FALSE}
# load the DESeq2 R package
library(DESeq2)
# create the DESeq2 main object
dds0 <- DESeqDataSetFromMatrix(countData = counts, colData = expDesign, design = ~ strain)
print(dds0)
```

## DESeq2 differential analysis with two command lines

```{r}
dds0 <- DESeq(dds0)
res0 <- results(dds0)
print(res0)
print(summary(res0))
```

## Sub-sampling: analysis using a few replicates

```{r}
nb.replicates <- 4
samples.WT <- sample(1:48, size=nb.replicates, replace=FALSE)
samples.Snf2 <- sample(49:96, size=nb.replicates, replace=FALSE)
print(samples.WT)
print(samples.Snf2)
dds <- DESeqDataSetFromMatrix(countData = counts[,c(samples.WT, samples.Snf2)], 
                              colData = expDesign[c(samples.WT, samples.Snf2),], 
                              design = ~ strain)
```

We now perform a differential analysis with DESeq2 step by step with some quality controls.

## Normalization

```{r fig.width=14}
dds <- estimateSizeFactors(dds)
print(sizeFactors(dds))
# effect of the normalization
par(mfrow=c(1,2))
boxplot(log2(counts(dds, normalized=FALSE)+1), main="Raw counts",
        col=rep(c("lightblue","orange"), each=nb.replicates))
boxplot(log2(counts(dds, normalized=TRUE)+1), main="Normalized counts",
        col=rep(c("lightblue","orange"), each=nb.replicates))
```

## Mean-variance relationship

```{r}
m <- rowMeans(counts(dds))
v <- apply(counts(dds), 1, var)
plot(m, v, pch=16, cex=0.3, main="Mean-variance relationship", log="xy")
abline(a=0, b=1)
```

We observe over-dispersion in the data: the Poisson distribution is not adapted.

## Principal Component Analysis (PCA)

```{r message=FALSE, fig.width=10}
# dispersions estimation
dds <- estimateDispersions(dds)
# make the data homoscedastic
rld <- rlog(dds) # alternative to the "Variance Stabilizing Transformation"
plotPCA(rld, intgroup="strain") # function from the DESeq2 package
```

## Statistical test for each gene

```{r}
dds <- nbinomWaldTest(dds)
res.DESeq2 <- results(dds, alpha=0.05, pAdjustMethod="BH")
print(head(res.DESeq2))
print(mcols(res.DESeq2))
```

## Statistical test for each gene (2)

```{r}
summary(res.DESeq2, alpha=0.05)
```

## Histogram of raw P-values

```{r}
hist(res.DESeq2$pvalue, col="lightblue", main="Histogram of raw P-values (DESeq2)", breaks=20, xlab="P-value")
```

## MA-plot

```{r}
plotMA(res.DESeq2, alpha=0.05) # function from the DESeq2 package
```

## Volcano-plot

Here we need to build the plot using R base functions:

```{r}
plot(x=res.DESeq2$log2FoldChange, y=-log10(res.DESeq2$padj), 
     xlab="log2(Fold-Change)", ylab="-log10(adjusted P-value",
     col=ifelse(res.DESeq2$padj<=0.05, "red", "black"), main="Volcano plot")
```

# Differential analysis using edgeR

## edgeR main object: dge

```{r message=FALSE}
# load the edgeR R package
library(edgeR)
# create the edgeR main object: dge
dge <- DGEList(counts=counts[,c(samples.WT, samples.Snf2)], remove.zeros=FALSE)
dge$design <- model.matrix(~ strain, data=expDesign[c(samples.WT, samples.Snf2),])
print(dge)
```

## Normalization & dispersions estimation

```{r}
# normalization
dge <- calcNormFactors(dge)
print(dge$samples$norm.factors)
# dispersions
dge <- estimateGLMCommonDisp(dge, dge$design)
dge <- estimateGLMTrendedDisp(dge, dge$design)
dge <- estimateGLMTagwiseDisp(dge, dge$design)
```

## Modeling and testing

```{r}
fit <- glmFit(dge, dge$design)
print(dge$design)
lrt <- glmLRT(fit, coef="strainSnf")
res.edgeR <- topTags(lrt,n=nrow(dge$counts),adjust.method="BH")$table
print(head(res.edgeR))
```

## Histogram of raw P-values

```{r}
hist(res.edgeR$PValue, col="lightblue", main="Histogram of raw P-values (edgeR)", breaks=20, xlab="P-value")
```

# Compare DESeq2 and edgeR results

## Normalization method

```{r}
plot(x=sizeFactors(dds), y=dge$samples$norm.factors,
     xlab="DESeq2 size factors", ylab="edgeR normalization factors", pch=16)
```

The normalization/size factors computed by DESeq2 and edgeR are not comparable as they are used in a different manner in the statistical/mathematical models.

## Re-order the results according to the gene names

```{r}
print(head(res.DESeq2))
print(head(res.edgeR))
res.edgeR <- res.edgeR[order(rownames(res.edgeR)),]
res.DESeq2 <- res.DESeq2[order(rownames(res.DESeq2)),]
```

## log2(Fold-Change) estimations

```{r}
plot(x=res.edgeR$logFC, y=res.DESeq2$log2FoldChange, 
     pch=16, cex=0.4, xlim=c(-2,2), ylim=c(-2,2),
     xlab="edgeR log2(Fold-Change)", ylab="DESeq2 log2(Fold-Change)")
abline(a=0, b=1, col="red", lwd=4) # draw the y=x curve (y=a+b*x with a=0 and b=1)
abline(h=0, v=0) # horizontal and vertical line
```

## Raw P-values

```{r}
plot(x=res.edgeR$PValue, y=res.DESeq2$pvalue, 
     pch=16, cex=0.4, xlab="edgeR raw P-value", ylab="DESeq2 raw P-value")
abline(a=0, b=1, col="red", lwd=4) # draw the y=x curve (y=a+b*x with a=0 and b=1)
```

## Number of differentially expressed genes

```{r}
# remember the number of replicates
print(nb.replicates)
# DESeq2
sum(res.DESeq2$padj <= 0.05, na.rm=TRUE)
# edgeR
sum(res.edgeR$FDR <= 0.05, na.rm=TRUE)
```

What's the behaviour of the number of differentially expressed genes according to the number of samples?

## Venn diagram

```{r}
library(gplots)
venn(list(DESeq2=rownames(res.DESeq2[which(res.DESeq2$padj <= 0.05),]), 
          edgeR=rownames(res.edgeR[which(res.edgeR$FDR <= 0.05),])))
```

Supplementary exercise: do the same plot for up- and down-regulated genes separately.

## What's edgeR or DESeq2-specific?

```{r}
DESeq2.genes <- rownames(res.DESeq2[which(res.DESeq2$padj <= 0.05),])
edgeR.genes <- rownames(res.edgeR[which(res.edgeR$FDR <= 0.05),])
# select a DESeq2 specific gene
spe.DESeq2 <- setdiff(DESeq2.genes, edgeR.genes)
summary(res.edgeR[spe.DESeq2,"FDR"])
# select a edgeR specific gene
spe.edgeR <- setdiff(edgeR.genes, DESeq2.genes)
summary(res.DESeq2[spe.edgeR,"padj"])
```

## edgeR vs DESeq2 results for one gene

```{r}
# plotCounts is a function from the DESeq2 R package
plotCounts(dds, gene="ycr017c", intgroup="strain", normalized=TRUE)
plotCounts(dds, gene="ycr021c", intgroup="strain", normalized=TRUE)
```

## Differential analysis under $H_0$

Here we perform a differential analysis in which we compare N WT samples vs N other WT samples.

```{r}
nb.replicates <- 10
samples.WT <- sample(1:48, size=2*nb.replicates, replace=FALSE)
print(samples.WT)
counts.H0 <- counts[,samples.WT]
expDesign.H0 <- expDesign[samples.WT,]
expDesign.H0$condition <- factor(rep(c("A","B"), each=nb.replicates))
print(expDesign.H0)
dds.H0 <- DESeqDataSetFromMatrix(countData = counts.H0, colData = expDesign.H0, design = ~ condition)
```

## Differential analysis with DESeq2 under $H_0$

```{r}
dds.H0 <- DESeq(dds.H0)
res.H0 <- results(dds.H0)
summary(res.H0)
```

# sessionInfo

Here is the list of the R packages used and their versions:

```{r}
sessionInfo()
```
