---
title: 'Practical: exploring RNA-Seq counts'
author: "Julie Aubert - Hugo Varet - Jacques van Helden"
date: '`r Sys.Date()`'
output:
  slidy_presentation:
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
---

## Context

Presentation of the experiment:

 - Data extracted from the publication: Gierliński et al. *Statistical models for RNA-seq data derived from a two-condition 48-replicate experiment*, Bioinformatics, 2015
 - Study of 48 WT yeast samples vs 48 Snf2 (KO) samples.

## Loading a data table

**R** enables to download data directly from the Web. 

Load counts per gene for each biological condition (WT and Snf2).

```{r}
## Load the files content in an R data.frame
data.path <- "http://jvanheld.github.io/stats_avec_RStudio_EBA/practicals/yeast_2x48_replicates/data"
Snf2 <- read.table(file=paste(data.path, "Snf2_raw.tsv", sep="/"), row.names=1)
WT <- read.table(file=paste(data.path, "WT_raw.tsv", sep="/"), row.names=1)
## look at the beginning of each dataset:
print(Snf2[1:4,1:5])
print(WT[1:4,1:5])
```

## Merging the two biological conditions (1)

```{r}
# dimension of the two tables (numbers of rows/columns)
dim(Snf2)
dim(WT)

# randomly select N samples for each condition
nb.replicates <- 4
WT <- WT[,sample(1:ncol(WT), nb.replicates, replace=FALSE)]
Snf2 <- Snf2[,sample(1:ncol(Snf2), nb.replicates, replace=FALSE)]

# redefine column names
colnames(Snf2) <- paste0("S", 1:ncol(Snf2))
colnames(WT) <- paste0("WT", 1:ncol(WT))
```

## Merging the two biological conditions (2)

```{r}
# check that the genes are in the same order
all(rownames(WT)==rownames(Snf2))

# combine the two tables
counts <- cbind(WT, Snf2)
print(counts[1:4,])
```

## Differential analysis using DESeq2

```{r message=FALSE}
# load DESeq2 package
library(DESeq2)
# create a table describing the experiment
expDesign <- data.frame(sample=colnames(counts), condition=ifelse(grepl("WT", colnames(counts)), "WT", "Snf"))
print(head(expDesign))
# create the DESeq2 main object: dds
dds <- DESeqDataSetFromMatrix(countData = counts, colData = expDesign, design = ~ condition)
print(dds)
```

## Basic description of the data (1)

```{r}
# number of reads per sample
barplot(colSums(counts)/1000000, main="Total number of reads per sample (million)")
```

## Basic description of the data (2)

```{r}
# proportion of null counts per sample
barplot(apply(counts, 2, function(x) 100*mean(x==0)), main="Percentage of null counts per sample")
```

## Normalization

```{r}
# normalization
dds <- estimateSizeFactors(dds)
print(sizeFactors(dds))
# effect of the normalization
normCounts <- counts(dds, normalized=TRUE)
par(mfrow=c(1,2))
boxplot(log2(counts+1), main="Raw counts")
boxplot(log2(normCounts+1), main="Normalized counts")
```

## Principal Component Analysis (PCA)

```{r message=FALSE}
# dispersions estimation
dds <- estimateDispersions(dds)
# Principal Component Analysis (PCA) plot
rld <- rlog(dds)
plotPCA(rld, intgroup="condition")
```

## Statistical test for each gene

```{r}
# statistical modeling and testing
dds <- nbinomWaldTest(dds)
res.DESeq2 <- results(dds, alpha=0.05, pAdjustMethod="BH")
print(head(res.DESeq2))
```

## Statistical test for each gene (2)

```{r}
summary(res.DESeq2, alpha=0.05)
```

## Histogram of raw P-values

```{r}
# histogram of raw P-values
hist(res.DESeq2$pvalue, col="lightblue", main="Histogram of raw P-values", breaks=20, xlab="P-value")
```

## MA-plot

```{r}
# MA-plot of the results
plotMA(res.DESeq2)
```

## Differential analysis using edgeR

```{r message=FALSE}
library(edgeR)
# create the edgeR main object: dge
dge <- DGEList(counts=counts, remove.zeros=FALSE)
dge$design <- model.matrix(~ condition, data=expDesign)
print(dge)
```

## edgeR normalization

```{r}
# normalization
dge <- calcNormFactors(dge)
print(dge$samples$norm.factors)
```

## Dispersions estimations

```{r}
# dispersions estimation
dge <- estimateGLMCommonDisp(dge, dge$design)
dge <- estimateGLMTrendedDisp(dge, dge$design)
dge <- estimateGLMTagwiseDisp(dge, dge$design)
```

## Modeling and testing

```{r}
# statistical modeling and testing
fit <- glmFit(dge, dge$design)
print(dge$design)
lrt <- glmLRT(fit, coef="conditionWT")
res.edgeR <- topTags(lrt,n=nrow(dge$counts),adjust.method="BH")$table
print(head(res.edgeR))
```

## Compare DESeq2 and edgeR results: normalization

```{r}
# normalization
plot(sizeFactors(dds), dge$samples$norm.factors, pch=16, xlab="DESeq2 size factors", ylab="edgeR normalization factors")
```

## Compare DESeq2 and edgeR results: log2FC estimations

```{r}
# log2FC estimations
plot(res.edgeR$logFC, res.DESeq2$log2FoldChange, pch=16, cex=0.4, xlim=c(-2,2), ylim=c(-2,2),
     xlab="edgeR log2(Fold-Change)", ylab="DESeq2 log2(Fold-Change)")
```

## Compare DESeq2 and edgeR results: reorder the results according to gene names

```{r}
# reorder the results according to gene names
res.edgeR <- res.edgeR[order(rownames(res.edgeR)),]
res.DESeq2 <- res.DESeq2[order(rownames(res.DESeq2)),]
plot(res.edgeR$logFC, res.DESeq2$log2FoldChange, pch=16, cex=0.4, xlim=c(-2,2), ylim=c(-2,2),
     xlab="edgeR log2(Fold-Change)", ylab="DESeq2 log2(Fold-Change)")
abline(a=0, b=1, col="red", lwd=4) # draw the y=x curve (y=a+b*x with a=0 and b=1)
abline(h=0, v=0)
```

## Compare DESeq2 and edgeR results: p-values

```{r}
plot(res.edgeR$PValue, res.DESeq2$pvalue, 
     pch=16, cex=0.4, xlab="edgeR raw P-value", ylab="DESeq2 raw P-value")
abline(a=0, b=1, col="red", lwd=4) # draw the y=x curve (y=a+b*x with a=0 and b=1)
```

## Number of differentially expressed genes

```{r}
# remember the number of replicates
print(nb.replicates)
# DESeq2
sum(res.DESeq2$padj <= 0.05, na.rm=TRUE)
# edgeR
sum(res.edgeR$FDR <= 0.05, na.rm=TRUE)
```
