---
title: 'Practical: exploring RNA-Seq counts'
author: "Julie Aubert - Hugo Varet - Jacques van Helden"
date: '`r Sys.Date()`'
output:
  slidy_presentation:
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
---

```{r include=FALSE}
options(width=300)
knitr::opts_chunk$set(fig.align="center", cache=FALSE)
```

## Context

 - Study of 48 WT yeast samples vs 48 Snf2 (KO) samples: Gierliński et al. *Statistical models for RNA-seq data derived from a two-condition 48-replicate experiment*, Bioinformatics, 2015
 - RNA-Seq reads have been cleaned, mapped and counted to generated a count data matrix

## Loading a data table

**R** enables to download data directly from the Web. Load counts per gene for each sample of biological condition (WT and Snf2).

```{r}
## Load the files content in an R data.frame
data.path <- "http://jvanheld.github.io/stats_avec_RStudio_EBA/practicals/yeast_2x48_replicates/data"
Snf2 <- read.table(file=paste(data.path, "Snf2_raw.tsv", sep="/"), header=FALSE, row.names=1)
WT <- read.table(file=paste(data.path, "WT_raw.tsv", sep="/"), header=FALSE, row.names=1)
## look at the beginning of each dataset:
print(Snf2[1:4,1:5])
print(WT[1:4,1:5])
```

## Merging the two matrices (1)

```{r}
# dimension of the two tables (numbers of rows/columns)
dim(Snf2)
dim(WT)

# randomly select N samples for each condition
nb.replicates <- 4
WT <- WT[,sample(1:ncol(WT), nb.replicates, replace=FALSE)]
Snf2 <- Snf2[,sample(1:ncol(Snf2), nb.replicates, replace=FALSE)]

# redefine column names
colnames(Snf2) <- paste("S", 1:ncol(Snf2), sep="")
colnames(WT) <- paste("WT", 1:ncol(WT), sep="")
```

## Merging the two matrices (2)

```{r}
# check that the genes are in the same order
all(rownames(WT)==rownames(Snf2))
# if TRUE then combine the matrices
counts <- cbind(WT, Snf2)
print(counts[1:5,])
```

**Goal**: find genes which are differentially expressed between the *N* WT and the *N* Snf2 samples (with *N* lower than 48!).

## Basic description of the data: number of reads per sample

```{r}
barplot(colSums(counts)/1000000, main="Total number of reads per sample (million)")
```

## Basic description of the data: percentage of null counts per sample

```{r}
barplot(apply(counts, 2, function(x) 100*mean(x==0)), main="Percentage of null counts per sample")
```

# Differential analysis using DESeq2

## DESeq2 main object: dds

```{r message=FALSE}
# load the DESeq2 R package
library(DESeq2)
# create a table describing the experiment
expDesign <- data.frame(sample=colnames(counts), condition=ifelse(grepl("WT", colnames(counts)), "WT", "Snf"))
print(head(expDesign))
# create the DESeq2 main object: dds
dds <- DESeqDataSetFromMatrix(countData = counts, colData = expDesign, design = ~ condition)
print(dds)
```

## Normalization

```{r fig.width=14}
dds <- estimateSizeFactors(dds)
print(sizeFactors(dds))
# effect of the normalization
normCounts <- counts(dds, normalized=TRUE)
par(mfrow=c(1,2))
boxplot(log2(counts+1), main="Raw counts")
boxplot(log2(normCounts+1), main="Normalized counts")
```

## Principal Component Analysis (PCA)

```{r message=FALSE, fig.width=10}
# dispersions estimation
dds <- estimateDispersions(dds)
# make the data homoscedastic
rld <- rlog(dds)
plotPCA(rld, intgroup="condition") # function from the DESeq2 package
```

## Statistical test for each gene

```{r}
dds <- nbinomWaldTest(dds)
res.DESeq2 <- results(dds, alpha=0.05, pAdjustMethod="BH")
print(head(res.DESeq2))
```

## Statistical test for each gene (2)

```{r}
summary(res.DESeq2, alpha=0.05)
```

## Histogram of raw P-values

```{r}
hist(res.DESeq2$pvalue, col="lightblue", main="Histogram of raw P-values", breaks=20, xlab="P-value")
```

## MA-plot

```{r}
plotMA(res.DESeq2) # function from the DESeq2 package
```

## Volcano-plot

Here we need to build the plot using R base functions:

```{r}
plot(x=res.DESeq2$log2FoldChange, y=-log10(res.DESeq2$padj), 
     xlab="log2(Fold-Change)", ylab="-log10(adjusted P-value",
     col=ifelse(res.DESeq2$padj<=0.05, "red", "black"), main="Volcano plot")
```

# Differential analysis using edgeR

## edgeR main object: dge

```{r message=FALSE}
# load the edgeR R package
library(edgeR)
# create the edgeR main object: dge
dge <- DGEList(counts=counts, remove.zeros=FALSE)
dge$design <- model.matrix(~ condition, data=expDesign)
print(dge)
```

## Normalization & dispersions estimation

```{r}
# normalization
dge <- calcNormFactors(dge)
print(dge$samples$norm.factors)
# dispersions
dge <- estimateGLMCommonDisp(dge, dge$design)
dge <- estimateGLMTrendedDisp(dge, dge$design)
dge <- estimateGLMTagwiseDisp(dge, dge$design)
```

## Modeling and testing

```{r}
fit <- glmFit(dge, dge$design)
print(dge$design)
lrt <- glmLRT(fit, coef="conditionWT")
res.edgeR <- topTags(lrt,n=nrow(dge$counts),adjust.method="BH")$table
print(head(res.edgeR))
```

# Compare DESeq2 and edgeR results

## Normalization method

```{r}
plot(x=sizeFactors(dds), y=dge$samples$norm.factors,
     xlab="DESeq2 size factors", ylab="edgeR normalization factors", pch=16)
```

The normalization/size factors computed by DESeq2 and edgeR are not comparable as they are used in a different manner in the statistical/mathematical models.

## Re-order the results according to the gene names

```{r}
print(head(res.DESeq2))
print(head(res.edgeR))
res.edgeR <- res.edgeR[order(rownames(res.edgeR)),]
res.DESeq2 <- res.DESeq2[order(rownames(res.DESeq2)),]
```

## log2(Fold-Change) estimations

```{r}
plot(x=res.edgeR$logFC, y=res.DESeq2$log2FoldChange, 
     pch=16, cex=0.4, xlim=c(-2,2), ylim=c(-2,2),
     xlab="edgeR log2(Fold-Change)", ylab="DESeq2 log2(Fold-Change)")
abline(a=0, b=1, col="red", lwd=4) # draw the y=x curve (y=a+b*x with a=0 and b=1)
abline(h=0, v=0) # horizontal and vertical line
```

## Raw P-values

```{r}
plot(x=res.edgeR$PValue, y=res.DESeq2$pvalue, 
     pch=16, cex=0.4, xlab="edgeR raw P-value", ylab="DESeq2 raw P-value")
abline(a=0, b=1, col="red", lwd=4) # draw the y=x curve (y=a+b*x with a=0 and b=1)
```

## Number of differentially expressed genes

```{r}
# remember the number of replicates
print(nb.replicates)
# DESeq2
sum(res.DESeq2$padj <= 0.05, na.rm=TRUE)
# edgeR
sum(res.edgeR$FDR <= 0.05, na.rm=TRUE)
```

What's the behaviour of the number of differentially expressed genes according to the number of samples?
