---
title: "Stats with R and RStudio"
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  slidy_presentation:
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    toc_float: yes
    widescreen: yes
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  word_document:
    toc: yes
    toc_depth: 3
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
subtitle: 'Practical 1: first steps with R'
address: TAGC lab, Aix-Marseille Université, France
transition: linear
---

```{r setup, include=FALSE, size="huge"}
library(knitr)
## Default parameters for displaying the slides
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, fig.width = 7, fig.height = 5, fig.align = "center", size = "tiny", warning = FALSE, results = TRUE, message = FALSE, comment = "")
```


## Scope

In this session we will explore basic manipulations of variables.

- Assigning a value to a variable
- Basic operations on numbers

## ***R*** is a calculator

**Convention:**

- Dark boxes: commands to type in RStudio **Console** (bottom-left panel).
- White boxes: the result you should obtain.

**Example:** compute a simple addition.  

```{r}
2 + 5
```
## Assign a value to a variable

In ***R*** **`<-`** means "create a variable and assign its value."

**Example:** 

- create a variable named $a$, 
- assign the value $2$ to this variable, 
- ***print*** the result. 

```{r}
a <- 2
print(a)
```

## Computing with variables

**Example:** 

- create a variable named $b$ having value $5$, 
- compute  $a + b$ and store the result in a variable named $c$,
- ***print*** the result. 

```{r}
b <- 5
c <- a + b
print(c)
```

## Variables need to be updated

**Example:** 

- change the value of $a$ to $3$, 
- print  the value of $c$
- Is this the correct result for $c = a + b$? Why?

```{r}
a <- 3 ## Change the value of a
print(c) ## Print the value of c

## Check whether c equals a + b
c == a + b
```

Note: **`==`** is used to test whether two variables have the same content. 

## Updating variable contents

**Example:** 

- change the value of $a$ to $27$, 
- recompute and print  the value of $c$

```{r}
a <- 27 ## Change the value of a
c <- a + b
print(c) ## Print the value of c

## Check whether c  equals a + b
c == a + b
```

## Vectors of values

The simplest data structure in ***R*** is a vector.
In the previous example, the variable ***a*** was actually a vector with a single value.

**Example:** create a variable named  ***three.numbers***, and initialize it with a vector with values ***27***, ***12*** and ***3000***.

**Tips:**
- variable names can comprize several parts, separated by dots.
- the function ***`c()`*** combines several values into a vector

```{r}
three.numbers <- c(27,12,3000)
print(three.numbers)
```


## Series

The simple way to create a series of numbers.
The column operator permits to generate all integer values between two limits.

```{r}
x <- 0:14
print(x)
```

## Computing with vectors


**R** handles vectors in a very convenient way. 
An operation on a vector applies to all its elements.

```{r}
x <- 1:10 # Define a series from 1 to 10
print(x)
y <- x^2 # Compute the square of each number
print(y)
```


## Scatter plot

```{r scatter_plot, fig.path="figures/", fig.width=7, fig.height=5, fig.align="center"}
x <- -10:10
y <- x^2
plot(x,y)
```

## Line plot


```{r line_plot, fig.path="figures/", fig.width=7, fig.height=5, fig.align="center"}
x <- -10:10
y <- x^2
plot(x,y, type="l")
```

## Variables can also contain strings

```{r}
# The # symbol allows to insert comments in R code

# Define  a vector named "whoami", and 
# containing two names
whoami <- c("Denis", "Siméon")
print(whoami) # Comment at the end of a line
```

## String concatenation

```{r}
# Define  a vector named "names", and 
# containing two names
whoami <- c("Denis", "Siméon")

# Paste the values of a vector of string 
print(paste(sep=" ", whoami[1], whoami[2]))
```


## Carl's preferred distribution

The function **`dpois()`** computes the Poisson **density**, i.e. the probability to observe **exactly** $x$ successes in a series of independent trials with equal probability.

The Poisson distribution is defined by a single parameter: the expected number of successes $\lambda$ (read "lambda").

$$P(X=x) = \frac{e^{-\lambda}\lambda^x}{x!}$$

```{r eval=FALSE}
x <- 0:14   # Define the X values from 0 to 14
y <- dpois(x, lambda = 2.5) # Poisson density
print(y) # Check the result
```

## Plotting the Poisson distribution


```{r dpois_plot, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
x <- 0:14   # Define the X values from 0 to 14
y <- dpois(x, lambda = 2.5) # Poisson density
plot(x,y) # Check the result
```

This first plot is not very nice. Let us get some help to improve it.

## Getting help for ***R*** functions


Need help? Type **`help()`**. 

```{r}
help(plot)
```

A question? Type `?`

```{r}
?plot
```

**Result:** ***R*** displays the help message for the function `dpois()`.

## Exercise: improve Poisson density plot


1. Do not (yet) look the next slide.
2. Read the help page for the `dpois()`function.
3. draw a plot that provides a didactic illustration of the  Poisson density.

## Improve the plot: type = histogram


```{r dpois_plot_h, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
x <- 0:14
lambda <- 2.54
y <- dpois(x, lambda)
plot(x,y, type="h")
```

## Improve the plot: Add a title


```{r dpois_plot_h_main, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
plot(x,y, type="h", lwd=5, col="blue",
     main="Poisson density")
```

## Improve the plot: define axis labels


```{r dpois_plot_h_xylab, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
plot(x,y, type="h", lwd=5, col="blue",
     main="Poisson density",
     xlab="x = number of successes",
     ylab="P(X=x)")
```

## Improve the plot: add a legend


```{r dpois_plot_h_legend, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
plot(x,y, type="h", lwd=5, col="blue",
     main="Poisson density",
     xlab="x = number of successes",
     ylab="P(X=x)")
legend("topright", paste("lambda =", lambda))
```


## Poisson: a family of curves


**Exercice:** explore the properties of the Poisson density function, by changing the rang of $x$ values, and the $\lambda$ parameter.



## Solution: a family of Poisson curves

$lambda=0.01$

```{r dpois_series_lambda0_01, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 0.01
x <- 0:20
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=5, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=0.1$

```{r dpois_series_lambda0_1, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 0.1
x <- 0:20
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=5, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=1$

```{r dpois_series_lambda1, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 1
x <- 0:20
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=5, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=2$

```{r dpois_series_lambda2, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 2
x <- 0:20
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=5, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=5$

```{r dpois_series_lambda5, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 5
x <- 0:20
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=5, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=5$

```{r dpois_series_lambda5_range40, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 5
x <- 0:40
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=5, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=10$

```{r dpois_series_lambda10, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 10
x <- 0:(4*lambda)
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=5, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=20$

```{r dpois_series_lambda20, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 20
x <- 0:(4*lambda)
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=2, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=30$

```{r dpois_series_lambda30, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 30
x <- 0:(2*lambda)
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=2, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=50$

```{r dpois_series_lambda50, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 50
x <- 0:(2*lambda)
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=2, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=100$

```{r dpois_series_lambda100, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 100
x <- 0:(2*lambda)
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=2, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=1000$

```{r dpois_series_lambda1000, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 1000
x <- 0:(2*lambda)
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=2, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=10000$

```{r dpois_series_lambda10000, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
lambda <- 10000
x <- 0:(2*lambda)
plot(x, dpois(x,lambda=lambda), type="h",   
     col="darkblue", lwd=2, xlab="X",ylab="dpois(x)")
legend("topright", paste("lambda=",lambda))
```

## Solution: a family of Poisson curves

$lambda=100000$

```{r dpois_series_lambda10000_rescaled, fig.path="figures/", fig.width=7, fig.height=4, fig.align="center"}
plot(95000:105000, dpois(95000:105000,lambda=100000), type="h", col="darkblue", xlab="X",ylab="dpois(x)")
```

## Before finishing -- keep track of your session

Tractability is an important issue in sciences. 
The ***R*** function `sessionInfo()` summarizes information about the versions of R, the operating system, and all the libraries used during a session. 

```{r}
sessionInfo()
```

