---
title: "Replicates, replicates, replicates, ..."
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 5
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 5
---


--------------------------------------------

## Abbreviations


| Abbrev | Description |
|--------|-------------|
| WT     | Wild-type   |
| Snf2   | Catalytic subunit of the SWI/SNF chromatin remodeling complex; involved in transcriptional regulation (source: [SGD](http://www.yeastgenome.org/locus/snf2/overview)) |
| SWI/SNF | SWItch/Sucrose Non-Fermentable: nucleosome remodeling complex found in both eukaryotes and prokaryotes (source: [Wikipedia](https://en.wikipedia.org/wiki/SWI/SNF)) |

--------------------------------------------

## Introduction

--------------------------------------------

## Data sources

- [Raw counts in WT](http://figshare.com/articles/Wild_type_yeast_gene_read_counts_from_48_replicate_experiment/1425503)

- [Raw counts in Snf2 mutant](http://figshare.com/articles/SNF2_knock_out_yeast_gene_read_counts_from_a_48_replicate_experiment/1425502)

- [Metadata (sample descriptions)](http://figshare.com/articles/Metadata_for_a_highly_replicated_two_condition_yeast_RNAseq_experiment_/1416210)

--------------------------------------------

## Loading data tables

```{r configuration}

## Define the path to download the data (I will replace this by an URL on github)
dir.data <- "~/Documents/enseignement/AVIESAN_ecole_bioinformatique/2015-09_Ecole_Bioinfo_AVIESAN_n4/stats_avec_RStudio_EBA15/data/yeast_48_replicates/"

list.files(dir.data)
```

### Raw count table for the wild-type (WT)

The function *read.table()* provides a flexible method to load data tables from text files with different formats. We can readily load a table indicating the raw counts for each yeast gene (row) in each replicate (column) of the WT. 

```{r data_loading_WT}

## Read the counts per gene in for the WT and Snf2 mutant, respectively
counts.WT.raw <- read.table(file.path(dir.data, "WT_raw.tsv"), row.names = 1, sep="\t")
head(counts.WT.raw) ## Display the first rows of the WT count table
dim(counts.WT.raw) ## Check the size of the WT count table

```


## Metadata

## Computing sample-wise (column-wise) statistics

Each column of the row count table corresponds to one replicate. Before analyzing the data, we would like to compute some global statistics for each sample. For this, *R* provides a convenient function called **summary()**. 

```{r summary, eval=TRUE, echo=TRUE}

summary(counts.WT.raw)

```


------------------------------------------------------------

## Exercise: Snf2 raw count table

1. Load the Snf2 raw count table from the file *Snf2_raw.tsv*, located in the same folder as the WT count table.
2. Count the number of genes and replicates.
3. Compute some summary statistics for each gene of the Snf2 count table: mean, variance, standard deviation, min, median, third quartile, max.
3; Compute the same summary statistics for each sample of the Snf2 count table.


### Solution

```{r data_loading_Snf2, eval=FALSE, echo=TRUE}

## Loading the Snf2 raw count table
counts.Snf2.raw <- read.table(file.path(dir.data, "Snf2_raw.tsv"), row.names = 1)
head(counts.WT.raw) ## Display the first rows of the Snf2 count table
dim(counts.WT.raw) ## Check the size of the Snf2 count table


## Sample-wise statistics
snf2.stats.per.sample <-  data.frame(
  mean = apply(counts.Snf2.raw, 2, mean),
  var = apply(counts.Snf2.raw, 2, var),
  sd = apply(counts.Snf2.raw, 2, sd),
  min = apply(counts.Snf2.raw, 2, min),
  median = apply(counts.Snf2.raw, 2, median),
  Q3 =  apply(counts.Snf2.raw, 2, quantile, 0.75),
  max = apply(counts.Snf2.raw, 2, max)
  )



## Gene-wise statistics
snf2.stats.per.gene <- data.frame(
  mean = apply(counts.Snf2.raw, 1, mean),
  var = apply(counts.Snf2.raw, 1, var),
  sd = apply(counts.Snf2.raw, 1, sd),
  min = apply(counts.Snf2.raw, 1, min),
  median = apply(counts.Snf2.raw, 1, median),
  Q3 =  apply(counts.Snf2.raw, 1, quantile, 0.75),
  max = apply(counts.Snf2.raw, 1, max)
  )


```

--------------------------------------------

## Acknowledgement

Thanks for Geoff Barton and Christian Cole for their instant answer to my Sundey night's, mail when I was searching for ready-to-use count tables for this course.

